Appendix to "New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2025-03-05 jso.qmd
small changes on tables

@fig-idx-proj-dist shows a sample of traces from the indexes in
@tbl-smoothness-squintability for the 6D sine wave data. This traces the
PP index values computed for projections along an interpolated path
between a randomly generated noise projection and the optimal projection (sine wave). (Note
index values were self-standardized to range between 0 and 1, for
comparison.) The `splines` index is very smooth with a gradual increase
in value between the noise projection and the optimal projection. The
`TIC` index is slightly less smooth and slightly less gradual change in
value. Least desirable is the `skinny` index which is noisier and the
optimization needs to be much closer to the optimal angle to have a
higher value. This illustrates how the new measures for squintability
and smoothness work to describe the PP indexes.

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: The columns $\nu$ and $\varsigma$ show the smoothness and squintability measures for three PP indexes for the sine wave data. The splines index has the largest squint angle and is the smoothest. The skinny index has low smoothness and smallest squint angle. As dimension increases the squint angle gets smaller but smoothness doesn't change much.
dt <- tibble(shape = c(rep("pipe", 5), rep("sine", 18)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = as.character(index),
         index = ifelse(index == "splines2d", "splines", index),
         variance = sqrt(variance)) 
dt |> 
  filter(shape == "sine", index %in% c("TIC", "splines", "skinny")) |>
  select(shape, index, d, smooth, squint) |>
  arrange(desc(squint)) |>
  knitr::kable(
    digits = 4, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      #"$\\eta$", "$\\ell$", 
      "$\\nu$", #"$\\sigma$",
      #"$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", 
      "$\\varsigma$"), linesep = "",
    booktabs = TRUE, escape = FALSE, row.names = FALSE, align = "|llr|rr") #align = "|llr|rrrr|rrrrr") |> 
  #column_spec(1, border_left = TRUE) |> 
  #column_spec(c(4, 8, 13), border_right = TRUE) |> 
  #kable_styling(font_size = 10)  
```

@@

@tbl-mod-output presents the results from fitting a logistic regression
model using the proportion of success as the response and smoothness,
squintability, dimension, number of jellyfish and maximum number of
iteration, as predictors. The fit suggests that the JSO success rate is only
affected by squintability, dimension and number of jellyfish. As expected,
the JSO success rate is higher with higher squintability and/or more
jellyfish, and lower when dimension is higher (higher-dimensional
optimization is more difficult). Interestingly, the JSO is unaffected by
the smoothness of the index function. This is consistent with the way
random search algorithms jump from value to value without taking local
regularity into account, as opposed to gradient-based optimizers for
example. Allowing for more iterations also does not affect success rate
significantly. A unit increase in squintability increases the success
rate by `r round(100 * (exp(broom::tidy(mod1)$estimate[3])-1), 0)`%. As
dimension increases by one the success rate almost halves. Increasing
the number of jellies by 10 increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[5])-1), 0)`%.
@tbl-joint-output shows the fit statistics supporting that the model is
reasonably strong.

@@

>>>>

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: The columns $\nu$ and $\varsigma$ show the smoothness and squintability measures for three PP indexes for the sine wave data. The splines index has the largest squint angle and is the smoothest. The skinny index has low smoothness and smallest squint angle. As dimension increases the squint angle gets smaller but smoothness doesn't change much.
dt <- tibble(shape = c(rep("pipe", 5), rep("sine", 18)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = as.character(index),
         index = ifelse(index == "splines2d", "splines", index),
         variance = sqrt(variance)) 
dt |> 
  filter(shape == "sine", index %in% c("TIC", "splines", "skinny")) |>
  mutate(index = factor(index, levels = c("splines", "TIC", "skinny"))) |> 
  select(shape, index, d, smooth, squint) |>
  arrange(index, d) |>
  knitr::kable(
    digits = 4, format = "latex", 
    col.names = c("shape", "index", "d", "$\\nu$", "$\\varsigma$"), linesep = c("", "", "\\hline"),
    booktabs = TRUE, escape = FALSE, row.names = FALSE, align = "|llr|rr") |> 
  #column_spec(1, border_left = TRUE) |>
  column_spec(c(3,5), border_right = TRUE) 
  #row_spec(c(3, 6), extra_latex_after = "\\cline{1-5}")
  #kable_styling(font_size = 10)   
```

@@

@@

@fig-idx-proj-dist shows a sample of traces from the indexes in
@tbl-smoothness-squintability for the 6D sine wave data. This traces the
PP index values computed for projections along an interpolated path
between a randomly generated noise projection and the optimal projection (sine wave). (Note
index values were self-standardized to range between 0 and 1, for
comparison.) The `splines` index is very smooth with a gradual increase
in value between the noise projection and the optimal projection. The
`TIC` index is slightly less smooth and slightly less gradual change in
value. Least desirable is the `skinny` index which is noisier and the
optimization needs to be much closer to the optimal angle to have a
higher value. This illustrates how the new measures for squintability
and smoothness work to describe the PP indexes.
@tbl-mod-output presents the results from fitting a logistic regression
model using the proportion of success as the response and smoothness,
squintability, dimension, number of jellyfish and maximum number of
iteration, as predictors. The fit suggests that the JSO success rate is only
affected by squintability, dimension and number of jellyfish. As expected,
the JSO success rate is higher with higher squintability and/or more
jellyfish, and lower when dimension is higher (higher-dimensional
optimization is more difficult). Interestingly, the JSO is unaffected by
the smoothness of the index function. This is consistent with the way
random search algorithms jump from value to value without taking local
regularity into account, as opposed to gradient-based optimizers for
example. Allowing for more iterations also does not affect success rate
significantly. A unit increase in squintability increases the success
rate by `r round(100 * (exp(broom::tidy(mod1)$estimate[3])-1), 0)`%. As
dimension increases by one the success rate almost halves. Increasing
the number of jellies by 10 increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[5])-1), 0)`%.
@tbl-joint-output shows the fit statistics supporting that the model is
reasonably strong.

### 2025-03-05 jso.tex
small changes on tables

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, colback=white, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, rightrule=.15mm, breakable, toprule=.15mm, leftrule=.75mm, opacityback=0, opacitybacktitle=0.6, toptitle=1mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, left=2mm]

@@

Figure~\ref{fig-idx-proj-dist} shows a sample of traces from the indexes
in Table~\ref{tbl-smoothness-squintability} for the 6D sine wave data.
This traces the PP index values computed for projections along an
interpolated path between a randomly generated noise projection and the
optimal projection (sine wave). (Note index values were
self-standardized to range between 0 and 1, for comparison.) The
\texttt{splines} index is very smooth with a gradual increase in value
between the noise projection and the optimal projection. The
\texttt{TIC} index is slightly less smooth and slightly less gradual
change in value. Least desirable is the \texttt{skinny} index which is
noisier and the optimization needs to be much closer to the optimal
angle to have a higher value. This illustrates how the new measures for
squintability and smoothness work to describe the PP indexes.

@@

\begin{tabular}{|llr|rr|llr|rr|llr|rr|llr|rr|llr|rr}
\toprule
shape & index & d & $\nu$ & $\varsigma$\\
\midrule
sine & splines & 4 & 3.5489 & 0.5723\\
sine & splines & 6 & 3.1149 & 0.5415\\
sine & splines & 8 & 2.7400 & 0.5394\\
sine & TIC & 4 & 3.4854 & 0.3839\\
sine & TIC & 8 & 2.7223 & 0.3785\\
sine & TIC & 6 & 3.0807 & 0.3747\\
sine & skinny & 6 & 2.2352 & 0.0973\\
sine & skinny & 4 & 2.5213 & 0.0810\\
\bottomrule
\end{tabular}

@@

Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using the proportion of success as the response and
smoothness, squintability, dimension, number of jellyfish and maximum
number of iteration, as predictors. The fit suggests that the JSO
success rate is only affected by squintability, dimension and number of
jellyfish. As expected, the JSO success rate is higher with higher
squintability and/or more jellyfish, and lower when dimension is higher
(higher-dimensional optimization is more difficult). Interestingly, the
JSO is unaffected by the smoothness of the index function. This is
consistent with the way random search algorithms jump from value to
value without taking local regularity into account, as opposed to
gradient-based optimizers for example. Allowing for more iterations also
does not affect success rate significantly. A unit increase in
squintability increases the success rate by 23\%. As dimension increases
by one the success rate almost halves. Increasing the number of jellies
by 10 increases the success rate by 32\%. Table~\ref{tbl-joint-output}
shows the fit statistics supporting that the model is reasonably strong.

@@

>>>>

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, arc=.35mm, opacitybacktitle=0.6, left=2mm, colback=white, bottomtitle=1mm, titlerule=0mm, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomrule=.15mm, rightrule=.15mm, coltitle=black, toprule=.15mm, breakable, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]

@@

@@

\begin{tabular}{|llr|rr|llr|rr|llr|rr|llr|rr|llr|rr}
\toprule
shape & index & d & $\nu$ & $\varsigma$\\
\midrule
sine & splines & 4 & 3.5489 & 0.5723\\
sine & splines & 6 & 3.1149 & 0.5415\\
sine & splines & 8 & 2.7400 & 0.5394\\
\hline
sine & TIC & 4 & 3.4854 & 0.3839\\
sine & TIC & 6 & 3.0807 & 0.3747\\
sine & TIC & 8 & 2.7223 & 0.3785\\
\hline
sine & skinny & 4 & 2.5213 & 0.0810\\
sine & skinny & 6 & 2.2352 & 0.0973\\
\bottomrule
\end{tabular}

@@

@@

Figure~\ref{fig-idx-proj-dist} shows a sample of traces from the indexes
in Table~\ref{tbl-smoothness-squintability} for the 6D sine wave data.
This traces the PP index values computed for projections along an
interpolated path between a randomly generated noise projection and the
optimal projection (sine wave). (Note index values were
self-standardized to range between 0 and 1, for comparison.) The
\texttt{splines} index is very smooth with a gradual increase in value
between the noise projection and the optimal projection. The
\texttt{TIC} index is slightly less smooth and slightly less gradual
change in value. Least desirable is the \texttt{skinny} index which is
noisier and the optimization needs to be much closer to the optimal
angle to have a higher value. This illustrates how the new measures for
squintability and smoothness work to describe the PP indexes.
Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using the proportion of success as the response and
smoothness, squintability, dimension, number of jellyfish and maximum
number of iteration, as predictors. The fit suggests that the JSO
success rate is only affected by squintability, dimension and number of
jellyfish. As expected, the JSO success rate is higher with higher
squintability and/or more jellyfish, and lower when dimension is higher
(higher-dimensional optimization is more difficult). Interestingly, the
JSO is unaffected by the smoothness of the index function. This is
consistent with the way random search algorithms jump from value to
value without taking local regularity into account, as opposed to
gradient-based optimizers for example. Allowing for more iterations also
does not affect success rate significantly. A unit increase in
squintability increases the success rate by 23\%. As dimension increases
by one the success rate almost halves. Increasing the number of jellies
by 10 increases the success rate by 32\%. Table~\ref{tbl-joint-output}
shows the fit statistics supporting that the model is reasonably strong.

### 2025-03-05 supplementary/supplementary.qmd
small changes on tables

# Gaussian process estimates for all PP indexes and data sets

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (outputscale $\eta$, lengthscale $\ell$, smoothness $\nu$, and nugget $\sigma$) and scaled logistic function ($\theta_1$ to $\theta_4$ and $\varsigma$) for the pipe-finding and sine-wave finding problems.  The columns $\nu$ and $\varsigma$ represent the smoothness and squintability measures respectively.
dt <- tibble(shape = c(rep("pipe", 5), rep("sine", 17)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = ifelse(index == "dcor2d_2", "dcor", index),
         index = ifelse(index == "loess2d", "loess", index),
         index = ifelse(index == "splines2d", "splines", index),
         variance = sqrt(variance)) 
dt |> 
  knitr::kable(
    digits = 2, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      "$\\eta$", "$\\ell$", "$\\nu$", "$\\sigma$",
      "$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", "$\\varsigma$"), linesep = "",
    booktabs = TRUE, escape = FALSE, row.names = FALSE, align = "|llr|rrrr|rrrrr") |> 
  #column_spec(1, border_left = TRUE) |> 
  #column_spec(c(4, 8, 13), border_right = TRUE) |> 
  kable_styling(font_size = 10) 
```

>>>>

# Calculated smoothness and squintability metrics for all PP indexes and data sets

Smoothness and squintability are metrics used to characterize the index 
function for PP tasks and are invariant to the optimizer and its 
hyper-parameters. @tbl-smoothness-squintability presents the Gaussian process
parameters (outputscale $\eta$, lengthscale $\ell$, smoothness $\nu$, and 
nugget $\sigma$) and scaled logistic function parameters ($\theta_1$ to 
$\theta_4$ and $\varsigma$) for all 23 PP problems investigated in our work. 
The columns $\nu$ and $\varsigma$ represent the smoothness and squintability 
measures respectively.

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (outputscale $\eta$, lengthscale $\ell$, smoothness $\nu$, and nugget $\sigma$) and scaled logistic function ($\theta_1$ to $\theta_4$ and $\varsigma$) for the pipe-finding and sine-wave finding problems.  The columns $\nu$ and $\varsigma$ represent the smoothness and squintability measures respectively.
dt <- tibble(shape = c(rep("pipe", 5), rep("sine", 18)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = as.character(index),
         index = ifelse(index == "dcor2d_2", "dcor", index),
         index = ifelse(index == "loess2d", "loess", index),
         index = ifelse(index == "splines2d", "splines", index),
         variance = sqrt(variance)) 
dt |> 
  knitr::kable(
    digits = 2, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      "$\\eta$", "$\\ell$", "$\\nu$", "$\\sigma$",
      "$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", "$\\varsigma$"), linesep = "",
    booktabs = TRUE, escape = FALSE, row.names = FALSE, align = "|llr|rrrr|rrrrr") |> 
  #column_spec(1, border_left = TRUE) |> 
  #column_spec(c(4, 8, 13), border_right = TRUE) |> 
  kable_styling(font_size = 10) 
```
